---
title: "Data Cleaning with R"
author: "Machocho Mengo"
date: "3/1/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reading the  Data
Loading the necessary packages

```{r, echo=TRUE, eval=TRUE}
library(dplyr)
```

Loading the data from the CSV file

```{r, echo=TRUE, eval=TRUE}
advertising <- read.csv("~/Downloads/advertising.csv")
adverts <- as.data.frame(advertising)
```

Previewing the first rows of the dataset
```{r, echo=TRUE, eval=TRUE}
head(adverts)
```

Previewing the last observations of the dataset
```{r, echo=TRUE, eval=TRUE}
tail(adverts)
```

Next, we check the structure of the dataset to see the datatypes and shape of the dataset. 
```{r, echo=TRUE, eval=TRUE}
str(adverts)
```
The output above showing the datatypes of the different columns. 
The dtype for the timestamp will be changed to datetime 
THe dtype for the City variable will be changed to string
The dataset has 10 columns and 1000 rows

## Data Cleaning

### COMPLETENESS
Checking for missing values
```{r, echo=TRUE, eval=TRUE}
is.na(adverts)
colSums(is.na(adverts))
```
This dataset has no missing values. It is complete. 

### CONSISTENCY
Duplicates
```{r, echo=TRUE, eval=TRUE}
duplicated(adverts)
```
This datset has no duplicates either! Seems like the dataset gods have been kind to us :-)

UNIQUE VALUES
```{r, echo=TRUE, eval=TRUE}
unique(adverts)
```

OUTLIERS
Let us first remind ourselves of the columns in our dataset. 
```{r, echo=TRUE, eval=TRUE} 
colnames(adverts, do.NULL = TRUE, prefix = "col")

```
BOXPLOT
We use boxplots to visualize the outliers
```{r, echo=TRUE, eval=TRUE}
num_cols <- adverts[, c("Daily.Time.Spent.on.Site", "Age", "Area.Income", "Daily.Internet.Usage")]
num_cols
boxplot(num_cols)
```
 
 Outliers are present in the Area.Income. We can investigate further using the IQR test to see how to handle the outliers. 
 
IQR
```{r, echo=TRUE, eval=TRUE}
area_income <- adverts[, c("Area.Income")]
area_income

iqr_AI <- IQR(area_income)
iqr_AI
```

Visualizing the column using boxplot
```{r, echo=TRUE, eval=TRUE}
boxplot(area_income)
```


Discarding the Outliers
```{r, echo=TRUE, eval=TRUE}
summary(area_income)
```
We will use the 1st quartile to set the benchmark for the minimum value. 

```{r, echo=TRUE, eval=TRUE}
benchmark <- 47032 - 1.5*18438.8325
benchmark
```
 Our benchmark value is set. Any value below the value will be considered an outlier and so it will be discarded from the dataset. 
 
```{r, echo=TRUE, eval=TRUE}
area_income <- area_income[area_income > benchmark]

summary(area_income)

boxplot(area_income)

```
The outliers have been effectively discarded.


### UNIFORMITY
Changing TimeStamp variable to Date type. We first split the variable into two columns: DATE and TIME
Then we apply the as.Date() function to change the datatypes to Date and as.Time()
Let us remind ourselves of the dataset
```{r, echo=TRUE, eval=TRUE}
head(adverts)
```

To load the package we need for the process of splitting the column and changing the data type.
```{r, echo=TRUE, eval=TRUE}
library(tidyr)
```
#### TIMESTAMP
Splitting the column into Date and Time
```{r, echo=TRUE, eval=TRUE}
adverts_2 <- separate(data = adverts, col = Timestamp, into = c("Date", "Time"), sep = " ")
head(adverts_2)
```
Let us check the datatypes of the new columns

```{r, echo=TRUE, eval=TRUE}
str(adverts_2)
```

Changing the datatype to Date and Time
```{r, echo=TRUE, eval=TRUE}
adverts_2$Date <- as.Date(adverts_2$Date, format = "%Y-%m-%d")

adverts_2$Time <- as.Date(adverts_2$Time, format = "%h-%m-$s")
```

#### MALE, CLICKED ON AD, CITY, COUNTRY AND AD TOPIC LINE
We change the first two variables into factor as they are categorical variables and the 3 into characters
```{r, echo=TRUE, eval=TRUE}

#changing the datatypes into factors
adverts_2$Male <- as.factor(adverts_2$Male)
adverts_2$Clicked.on.Ad <- as.factor(adverts_2$Clicked.on.Ad)

#changing datatypes into characters
adverts_2$City <- as.character(adverts_2$City)
adverts_2$Country <- as.character(adverts_2$Country)
adverts_2$Ad.Topic.Line <- as.character(adverts_2$Ad.Topic.Line)

str(adverts_2)
```


## UNIVARIATE ANALYSIS
This is the analysis and visualization of individual variables of a dataset. 

### Daily.Time.Spent.on.Site
First, let us get the measures of dispersion and central tendencies

#### Measures of Central Tendencies
```{r, echo=TRUE, eval=TRUE}
#getting the artithmetic mean of the variable
mean_1 <- mean(adverts_2$Daily.Time.Spent.on.Site)
mean_1

#getting the median of the variable
meadian_1 <- median(adverts_2$Daily.Time.Spent.on.Site)
meadian_1

#creating a fxn for getting the mode 
getmode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}
   
mode_1 <- getmode(adverts_2$Daily.Time.Spent.on.Site)
mode_1
```

#### Measures of Dispersion
```{r, echo=TRUE, eval=TRUE}
#Getting the minimum value of the variable
min_1 <- min(adverts_2$Daily.Time.Spent.on.Site)
min_1

#getting the maximum value of the variable
max_1 <- max(adverts_2$Daily.Time.Spent.on.Site)
max_1

#Getting thw range of the variable
range_1 <- range(adverts_2$Daily.Time.Spent.on.Site)
range_1

#getting the quantiles of the variable 
quantiles_1 <- quantile(adverts_2$Daily.Time.Spent.on.Site)
quantiles_1

#getting the  variable of the varibale
var_1 <- var(adverts_2$Daily.Time.Spent.on.Site)
var_1

#getting the standard deviation of the variable 
std_1 <- sd(adverts_2$Daily.Time.Spent.on.Site)
std_1
```

#### Visualizations 

```{r, echo=TRUE, eval=TRUE}
#vizualization using boxplot
boxplot(adverts_2$Daily.Time.Spent.on.Site)

#visualization using the bar graphs approach
barplot(adverts_2$Daily.Time.Spent.on.Site)

#visualization using histograms
hist(adverts_2$Daily.Time.Spent.on.Site)
```

### AGE

#### Measures of Central Tendencies
```{r, echo=TRUE, eval=TRUE}
#getting the artithmetic mean of the variable
mean_2 <- mean(adverts_2$Age)
mean_2

#getting the median of the variable
meadian_2 <- median(adverts_2$Age)
meadian_2

#creating a fxn for getting the mode 
getmode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}
   
mode_2 <- getmode(adverts_2$Age)
mode_2
```

#### Measures of Dispersion
```{r, echo=TRUE, eval=TRUE}
#Getting the minimum value of the variable
min_2 <- min(adverts_2$Age)
min_2

#getting the maximum value of the variable
max_2 <- max(adverts_2$Age)
max_2

#Getting thw range of the variable
range_2 <- range(adverts_2$Age)
range_2

#getting the quantiles of the variable 
quantiles_2 <- quantile(adverts_2$Age)
quantiles_2

#getting the  variable of the varibale
var_2 <- var(adverts_2$Age)
var_2

#getting the standard deviation of the variable 
std_2 <- sd(adverts_2$Age)
std_2
```

#### Visualizations 

```{r, echo=TRUE, eval=TRUE}
#vizualization using boxplot
boxplot(adverts_2$Age)

#visualization using the bar graphs approach
barplot(adverts_2$Age)

#visualization using histograms
hist(adverts_2$Age)
```

### AREA.INCOME

#### Measures of Central Tendencies
```{r, echo=TRUE, eval=TRUE}
#getting the artithmetic mean of the variable
mean_3 <- mean(adverts_2$Area.Income)
mean_3

#getting the median of the variable
meadian_3 <- median(adverts_2$Area.Income)
meadian_3

#creating a fxn for getting the mode 
getmode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}
   
mode_3 <- getmode(adverts_2$Area.Income)
mode_3
```

#### Measures of Dispersion
```{r, echo=TRUE, eval=TRUE}
#Getting the minimum value of the variable
min_3 <- min(adverts_2$Area.Income)
min_3

#getting the maximum value of the variable
max_3 <- max(adverts_2$Area.Income)
max_3

#Getting thw range of the variable
range_3 <- range(adverts_2$Area.Income)
range_3

#getting the quantiles of the variable 
quantiles_3 <- quantile(adverts_2$Area.Income)
quantiles_3

#getting the  variable of the varibale
var_3 <- var(adverts_2$Area.Income)
var_3

#getting the standard deviation of the variable 
std_3 <- sd(adverts_2$Area.Income)
std_3
```

#### Visualizations 

```{r, echo=TRUE, eval=TRUE}
#vizualization using boxplot
boxplot(adverts_2$Area.Income)

#visualization using the bar graphs approach
barplot(adverts_2$Area.Income)

#visualization using histograms
hist(adverts_2$Area.Income)
```

### Daily.Internet.Usage

#### Measures of Central Tendencies
```{r, echo=TRUE, eval=TRUE}
#getting the artithmetic mean of the variable
mean_4 <- mean(adverts_2$Daily.Internet.Usage)
mean_4

#getting the median of the variable
meadian_4 <- median(adverts_2$Daily.Internet.Usage)
meadian_4

#creating a fxn for getting the mode 
getmode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}
   
mode_4 <- getmode(adverts_2$Daily.Internet.Usage)
mode_4
```

#### Measures of Dispersion
```{r, echo=TRUE, eval=TRUE}
#Getting the minimum value of the variable
min_4 <- min(adverts_2$Daily.Internet.Usage)
min_4

#getting the maximum value of the variable
max_4 <- max(adverts_2$Daily.Internet.Usage)
max_4

#Getting thw range of the variable
range_4 <- range(adverts_2$Daily.Internet.Usage)
range_4

#getting the quantiles of the variable 
quantiles_4 <- quantile(adverts_2$Daily.Internet.Usage)
quantiles_4

#getting the  variable of the varibale
var_4 <- var(adverts_2$Daily.Internet.Usage)
var_4

#getting the standard deviation of the variable 
std_4 <- sd(adverts_2$Daily.Internet.Usage)
std_4
```

#### Visualizations 

```{r, echo=TRUE, eval=TRUE}
#vizualization using boxplot
boxplot(adverts_2$Daily.Internet.Usage)

#visualization using the bar graphs approach
barplot(adverts_2$Daily.Internet.Usage)

#visualization using histograms
hist(adverts_2$Daily.Internet.Usage)
```

#### Visualizing the Male and Clicked on Ads variables. 

```{r, echo=TRUE, eval=TRUE}

#histogram of the Male variable 
hist(adverts$Male)

#histogram of the Clicked on Ads variable
boxplot(adverts$Clicked.on.Ad)

```

## BIVARIATE ANALYSIS

### Daily.Time.Spent.on.Site AND Age

```{r, echo=TRUE, eval=TRUE}
Time_s <- adverts_2$Daily.Time.Spent.on.Site
age <- adverts_2$Age

#getting the covariance between the two variables
cov(Time_s, age )

#getting the ccorrelation coefficient 
cor(Time_s, age)

#The variables have a negative correlaton coefficient. This means that for an increase in age, there is a -0.3315133 decrease in the time that one spends on the internet. The value is 0.3 which means that the correlation is very small 
```

SCATTERPLOTS
A scatter plot is a two-dimensional data visualization that uses dots to represent the values obtained for two different variables - one plotted along the x-axis and the other plotted along the y-axis. Scatter plots are used when you want to show the relationship between two variables. They are sometimes called correlation plots because they show how two variables are correlated.

```{r, echo=TRUE, eval=TRUE}

plot(Time_s, age, xlab="Time speent on the internet", ylab="Age of user")

plot(age, Time_s, xlab="Age of user", ylab="Time spent on the internet")

```

### AREA INCOME AND DAILY TIME SPENT ON THE SITE

```{r, echo=TRUE, eval=TRUE}
Time_s <- adverts_2$Daily.Time.Spent.on.Site
income <- adverts_2$Area.Income

#getting the covariance between the two variables
cov(Time_s, income )

#getting the ccorrelation coefficient 
cor(Time_s, income)

#there is a positive correlation between the 2 variables. This means that for every unite increase of income made, there is a 0.3109544 increase in time spent on the site. The correlation is weak howerver since the correlation coeficient is less than 0.5
```

SCATTERPLOT
```{r, echo=TRUE, eval=TRUE}

plot(Time_s, income, xlab="Time speent on the internet", ylab="Income of user")
```

### AREA INCOME AND AGE

```{r, echo=TRUE, eval=TRUE}
#getting the covariance between the two variables
cov(age, income )

#getting the ccorrelation coefficient 
cor(age, income)

#there is a very weak negative correlation between the two variables
```

SCATTERPLOT
```{r, echo=TRUE, eval=TRUE}

plot(age, income, xlab="Age of the user", ylab="Income of user")
```

### AGE AND DAILY INTERNET USAGE

```{r, echo=TRUE, eval=TRUE}
Daily <- adverts_2$Daily.Internet.Usage

#getting the covaraiance between the 2 variables
cov(age, Daily)

#getting the correlation coefficient
cor(age, Daily)

```
SCATTERPLOT
```{r, echo=TRUE, eval=TRUE}

plot(age, Daily, xlab="Age of the user", ylab="Income of user")
```

### INCOME AND DAILY INTERNET USAGE

```{r, echo=TRUE, eval=TRUE}

#getting the covaraiance between the 2 variables
cov(income, Daily)

#getting the correlation coefficient
cor(income, Daily)

```

SCATTERPLOT
```{r, echo=TRUE, eval=TRUE}

plot(Daily, income, xlab="Daily internet usage", ylab="Income of user")
```

### DAILY TIME SPENT ON SITE AND DAILY INTERNET USAGE

```{r, echo=TRUE, eval=TRUE}
#getting the covaraiance between the 2 variables
cov(Time_s, Daily)

#getting the correlation coefficient
cor(Time_s, Daily)

#There is a strong positive correlation between the 2 variables. Which means that for every unit increase in time spent on the internet, there is a 0.5186585 increase in time spent on the site. 
```

SCATTERPLOT
```{r, echo=TRUE, eval=TRUE}

plot(Time_s, Daily, xlab="Time spent on the site", ylab="Daily internet usage")
```
### CLICKED ON ADS VISUALIZATIONS

# Grouped Bar Plot
```{r, echo=TRUE, eval=TRUE}
counts = table(adverts$Male, adverts$Clicked.on.Ad)
barplot(counts, main="number of Clicks on an Ad as per each sex, 0=Female, 1=male",
xlab="Ad Clicks", col=c("green","red"),
legend = rownames(counts), beside=TRUE)
```
There is an equal number of both male and female users who did not click on ads. 
MOre females clicked on ads compared to men.

# Grouped Bar Plot
```{r, echo=TRUE, eval=TRUE}

counts = table(adverts$Age, adverts$Clicked.on.Ad)
barplot(counts, main="number of Clicks on an Ad as per Age",
xlab="Ad Clicks", col=c("yellow","blue"),
legend = rownames(counts), beside=TRUE)
```

## MULTIVARIATE ANALYSIS
> Getting the correlation matrix 
> Visualizing the correlation matrix
> Variable reduction  using the PCA approach. 

###CORRELATION MATRIX
We first select all the numeric variables from the dataframe
Then from the variables, we find the correlation matrix
```{r, echo=TRUE, eval=TRUE}
adverts_2.cor <- cor(num_cols, method = c("spearman"))
adverts_2.cor
```

```{r, echo=TRUE, eval=TRUE}
install.packages("Hmisc")
```

```{r, echo=TRUE, eval=TRUE}
install.packages("corrplot")
```

```{r, echo=TRUE, eval=TRUE}
library("corrplot")
corrplot(adverts_2.cor, method="color")
```

### VARIABLE REDUCTION USING THE PRINCIPAL COMPONENT ANALYSIS

We really do not need to reduce the variables for this dataset as we are not going to go further with modelling. We shall however cover that as a way of introducing ourselves to PCA in R

```{r, echo=TRUE, eval=TRUE}

adverts.pca <- prcomp(num_cols, center = TRUE,scale. = TRUE)
adverts.pca

```

# PUBLISHING THE WORK
```{r, echo=FALSE, eval=FALSE}
install.packages("RCurl")
install.packages("openssl")
install.packages("rsconnect")
```

```{r, echo=FALSE, eval=FALSE}
library("RCurl")
library("openssl")
library("rsconnect")
```



